# Central tendency

```{r setup, include=FALSE, echo=FALSE}
suppressMessages(library("tidyverse"))
```


Measures of central tendency is a fancy phrase for 'average'.
They are a single data point used to represent a 'typical' value from your data.
Depending on your level of measurement you can use one or more measures of central tendency.


## Mode

The most common unique value.
For example, 10 people are asked their favourite drink and you have the following results:

```{r drinks, include=TRUE, echo=FALSE}
drinks = tibble(
  choice = c(
    "espresso", "latte", "tea", "tea", "earl gray", "americano", 
    "hot chocolate", "tea", "hot chocolate", "tea"
  )
)

drinks$choice
```

The most common result (mode) is `tea`, which we can confirm with a quick frequency table:

```{r table-drinks}
drinks %>% 
  count(choice) %>% 
  arrange(desc(n))
```

Mode is the only measure of central tendency you can provide for [nominal data](../levels-of-measurement#nominal).


## Median

The median is the 'middle' point.
It's only appropriate for ordered data and is calculated by arranging your data in order and selecting the mid--point.
Consider the following incomes:

```{r median, include=TRUE, echo=FALSE}
set.seed(42)
dat = tibble(
  income = qnorm(seq(0, 1, 0.02), 2.73, 1) * 10000
) %>% 
  filter(income > 0, income != Inf)

as.vector(dat$income)
```

If we arrange these in order from lowest to highest we get:

```{r median-arranged, include=TRUE, echo=FALSE}
dat %>% 
  arrange(income) %>%
  unlist() %>% 
  as.vector()
```

The median is the middle point (note that this is not realistic data):

```{r median-result, include=TRUE, echo=FALSE}
median(dat$income)
```
