# Measures of spread

We saw in the measures of central tendency chapter that the mean can be a poor representation of data if the data is skewed, and that we should therefore be careful when someone presents us with a mean (or average) without any further information.

One of the types of 'further information' that can help us is a measure of spread of the data around the mean value.
We usually use the *variance* and the *standard deviation* to quantify measure of spread.
Both are easy to calculate, and even easier to convert between each other.


## Variance

Lets recap: this is what the income data looks like:

```{r income}
dat %>% 
  unlist() %>% 
  as.vector()
```

To calculate the variance:

1. subtract the mean from each score
1. square the result
1. sum the results to produce one value
1. divide by $n - 1$ (number of observations minus one)

$$
(\#eq:variance)
\frac{\Sigma (x - \bar{x}) ^ 2}{N-1}
$$

Using $n - 1$ rather than simply the number of observations is known as [Bessel's correction](https://en.wikipedia.org/wiki/Bessel%27s_correction).

```{r income-variance}
var(dat$income)
```


## Standard deviation

As you've probably noticed the variance is not in the units of the original data (otherwise the variance would be £`r var(dat$income)`).
This is where the standard deviation comes in.
In fact the unit of the variance is the *square* of the unit of the original data.
The standard deviation is therefore a measure of spread in the unit of the original data, and is calculated simply by square rooting the variance.

$$
(\#eq:standard-deviation)
\sqrt{\frac{\Sigma (x - \bar{x}) ^ 2}{N-1}}
$$

The standard deviation of the income is therefore:

```{r income-sd}
sd(dat$income)
```

The standard deviation is a measure of how far the data points are on average from the mean.
A small standard deviation means the mean fairly accurately represents the data; a large standard deviation means the mean does not represent the data well.
